---
title: "Proteomics differently expressed analysis between adenocarcinoma and healthy/inflammatory tissue"
output: html_document
date: '2022-03-29'
author: 'Pan Zhang'
---

## Description

#### Proteomics differently expressed analysis between adenocarcinoma (ADK) and healthy (DH) / inflammatory (DI) tissue using data in paper Quesada-Calvo et al 2017

## Procedure

#### Step 1. Download the data from https://clinicalproteomicsjournal.biomedcentral.com/articles/10.1186/s12014-017-9143-3#Sec19

#### Step 2. Check normality of data by histogram & Shapiro-Wilk Normality Test -> doesn't follow normal distribution

#### Step 3. PCA plot to check the similarity between and within each group -> samples from healthy and inflammatory tissue are gathered

#### Step 4. Differential test for 3 contrast (ADK vs DH, ADK vs DI, ADK vs DI&DH) by ANOVA and Kruskal-Wallis Rank Sum Test, calculate log2 fold change and metric (modified fold change)

#### Step 5. Venn plot for the result of 3 contrast (p Value < 0.05 & metric > 1)




```{r setup, include=FALSE}
library(lattice)
library(ggfortify)
library(gridExtra)
library(VennDiagram)
```


```{r, include=FALSE}
cutoff_perc <- function(x, perc = 25){
  sort_x <- sort(x)
  v <- sort_x[round(length(x)*perc*0.01 )]
  return (as.numeric(v))
}


cal_metric <- function(x, cutoff){
#Input: a list of expression value with 4 or 8 control and 8 adenocarcinoma  
#Output: modified fold change of cancer versus control  
#metric= ((Trimmed Fold change)*CancerProportion)/(ControlProportion+0.01)

    
  if (length(x) == 12){
    n_control = 4
    n_cancer = 8
  }else{
    n_control = 8
    n_cancer = 8
  }
  
  if (mean(as.numeric(x[(n_control+1):(n_control + n_cancer)])) != 0 & mean(as.numeric(x[1:n_control])) != 0 ){
    fc <- mean(as.numeric(x[(n_control+1):(n_control + n_cancer)])) /mean(as.numeric(x[1:n_control]))
  }else{
    fc <- mean(as.numeric(x[(n_control+1):(n_control + n_cancer)])+0.01) /mean(as.numeric(x[1:n_control])+0.01)
  }
  
  ControlProp<- as.numeric(table(as.numeric(x[1:n_control])> cutoff[1:n_control])['TRUE'])/n_control
  
  CancerProp <- as.numeric(table(as.numeric(x[(n_control+1):(n_control + n_cancer)])> cutoff[(n_control+1):(n_control + n_cancer)])['TRUE'])/n_cancer
  
  if (is.na(ControlProp)){
    metric <- fc*CancerProp/0.0001
  }
  else{
    metric <- fc*CancerProp/ControlProp
  }
  
  #return(list(log2(lfc),metric,ControlProp,CancerProp ))
  return(metric)
}


cal_logfc <- function(x){
#Input: a list of expression value with 4 or 8 control and 8 adenocarcinoma  
#Output: log 2 fold change of adenocarcinoma versus control  
#logFC = log 2 (mean(cancer)/mean(control)), if mean(cancer) ==0 or mean(control) ==0, add 0.01 for both
  if (length(x) == 12){
    n_control = 4
    n_cancer = 8
  }
  else{
    n_control = 8
    n_cancer = 8
  }
  
  if (mean(as.numeric(x[(n_control+1):(n_control + n_cancer)])) != 0 & mean(as.numeric(x[1:n_control])) != 0 ){
    lfc <- log2(mean(as.numeric(x[(n_control+1):(n_control + n_cancer)])) /mean(as.numeric(x[1:n_control])))
  }
  else {
    lfc <- log2(mean(as.numeric(x[(n_control+1):(n_control + n_cancer)])+0.01) /mean(as.numeric(x[1:n_control])+0.01))
  }
  return(lfc)
}

```

```{r, include=FALSE}
table_ouput<- function(val_contrast,mydata,cof){
    if (dim(val_contrast)[1] == 12){
    n_control = 4
    n_cancer = 8
  }
  else{
    n_control = 8
    n_cancer = 8
  }
  # filtering protein expressed in <2 samples
  keep <- colSums(val_contrast>0) >= 2
  colnames(val_contrast) <- mydata$Accession
  val_contrast <-val_contrast[,keep]
  dim(val_contrast)
  
  
  
  lfcs <- apply(val_contrast,2, cal_logfc )
  metrics <- apply(val_contrast,2, cal_metric, cutoff=cof )

  group<- factor(c(rep("DH",n_control),rep("ADK",n_cancer)))
  
  #Log transformation: log2(value + 0.01)
  logVal<-log2(val_contrast +0.01)
  colnames(logVal) <- colnames(val_contrast)

  Pvalue_anova<-c()
  Pvalue_ks <- c()
  for(i in 1:(ncol(logVal))){
    y <- aov(as.numeric(logVal[,i]) ~ group)
    Pvalue_anova[i]<-summary(y)[[1]][,5][1]
    ks <- kruskal.test(as.numeric(logVal[,i]) ~ group)
    Pvalue_ks[i] <- ks$p.value
  }

  data.exp <- t(val_contrast)
  acc <- mydata$Accession[keep]
  desc <- mydata$Description[keep]


  foo <- data.frame(do.call('rbind', strsplit(as.character(desc),'GN=|PE=')))

  Gene <- foo$X2


  data.exp <- cbind(data.exp, lfcs, metrics, Pvalue_anova, Pvalue_ks, Gene , desc)
  return(as.data.frame(data.exp))
  
}
```




```{r}
setwd("/Users/panzhang/Desktop/GGD_work/CRC/CRC_paper/proteomics_adenomas3")
mydata <- read.delim("CRC_proteomics_adenomas3.txt")
dim(mydata)
```

# Normality test & visualization 
```{r}
group_all <- factor(c(rep("DH",4),rep("DI",4),rep("ADK_T1",4 ),rep("ADK_T2",4 )))
expr <- mydata[apply(mydata[,12:27], 1, var, na.rm = T) != 0, ]
rownames(expr) <- expr$Accession
expr <- expr[,12:27]
dim(expr)
nortest <- c()
nortest_log <- c()

nortest_r0 <- c()
nortest_log_r0 <- c()

#Log transformation: log2(value + 0.01)
expressed <- c()
par(mfcol=c(2,2))
for (i in 1:16){
  nortest_log<- append(nortest_log,   shapiro.test(log2(expr[,i]+0.01))$p.value)
  nortest<- append(nortest, shapiro.test(expr[,i])$p.value)
  hist(log2(expr[,i]+0.01), breaks = 100, freq=FALSE)
  
  #remove 0s
  nortest_log_r0<- append(nortest_log_r0,   shapiro.test(log2(expr[expr[,i]>0,][,i]))$p.value)
  nortest_r0<- append(nortest_r0, shapiro.test(expr[expr[,i]>0,][,i])$p.value)
  
  hist(log2(expr[expr[,i]>0,][,i]), breaks = 100, freq=FALSE)
  #lines(density(log2(expr[,i]+0.01)),col="blue")
  expressed <- append(expressed, as.numeric(table(expr[,i] >0)["TRUE"])/length(expr[,i]))
  
}
colnames(expr)
nortest_log
nortest

nortest_log_r0
nortest_r0
expressed
par(mfcol=c(1,1))

barplot(expressed, col = "light blue", names.arg = colnames(expr))
```


# PCA plot 
### PC1 vs PC2
```{r}
expr.pca <- t(expr)
dim(expr.pca)

#PCA is based on original abundance after removing protein doesn't express across all samples
expr.pca <- expr.pca[, apply(expr.pca, 2, var, na.rm = T) != 0]
dim(expr.pca)
expr.pca <- prcomp(expr.pca, center = T, scale. = T)

expr_plot<-as.data.frame(t(expr))
expr_plot["group_all"] <- group_all
dim(expr_plot)
par(mfcol=c(2,2))
plot1 <- autoplot(expr.pca, data = expr_plot,colour = "group_all")
plot2 <- autoplot(expr.pca, data = expr_plot,colour = "group_all", label.size = 3, label = TRUE, )
#grid.arrange(plot1, plot2, nrow=2, ncol=1, heights=c(2,2))
plot1
plot2
```

# DE test
## ADK VS DI
```{r}
DI<- as.data.frame(t(cbind(mydata[,16:19], mydata[,20:27])))
cof_DI <- apply(DI, 1, cutoff_perc)
cof_DI
ADK_VS_DI <- table_ouput(DI, mydata, cof_DI)

colnames(ADK_VS_DI)[13:18] <-c("log2FC (ADK vs DI)", "Metric", "Pvalue anova", "Pvalue Kruskal", "Genes","Description" )
dim(ADK_VS_DI)

#write.table(ADK_VS_DI, "CRC_proteomics_adenomas3_ADK_VS_DI.txt", quote = F, sep = "\t")
```


## ADK VS DH
```{r}
DH <- as.data.frame(t(cbind(mydata[,12:15], mydata[,20:27])))
cof_DH <- apply(DH, 1, cutoff_perc)
cof_DH
ADK_VS_DH <- table_ouput(DH, mydata,cof_DH)

colnames(ADK_VS_DH)[13:18] <-c("log2FC (ADK vs DH)", "Metric", "Pvalue anova","Pvalue Kruskal", "Genes","Description" )
dim(ADK_VS_DH)

#write.table(ADK_VS_DH, "CRC_proteomics_adenomas3_ADK_VS_DH.txt", quote = F, sep = "\t")
```

## ADK vs DI+DH 
```{r}
DHDI<- as.data.frame(t(cbind(mydata[,12:19], mydata[,20:27])))
cof_DHDI <- apply(DHDI, 1, cutoff_perc)
cof_DHDI
ADK_VS_DHDI <- table_ouput(DHDI, mydata,cof_DHDI)
colnames(ADK_VS_DHDI)[17:22] <-c("log2FC (ADK vs DH&DI )", "Metric", "Pvalue anova","Pvalue Kruskal", "Genes", "Description" )
dim(ADK_VS_DHDI)

#write.table(ADK_VS_DHDI, "CRC_proteomics_adenomas3_ADK_VS_DH&DI.txt", quote = F, sep = "\t")

```

# Compare 3 contrast
## ANOVA
```{r}
#gene_DHDI <- ADK_VS_DHDI[which(ADK_VS_DHDI$`Pvalue anova`<0.05), ]$Genes
#gene_DH <- ADK_VS_DH[which(ADK_VS_DH$`Pvalue anova`<0.05), ]$Genes
#gene_DI <- ADK_VS_DH[which(ADK_VS_DI$`Pvalue anova`<0.05), ]$Genes

gene_DHDI <- ADK_VS_DHDI[which(ADK_VS_DHDI$`Pvalue anova`<0.05 & as.numeric(ADK_VS_DHDI$Metric) >1), ]$Genes
gene_DH <- ADK_VS_DH[which(ADK_VS_DH$`Pvalue anova`<0.05 & as.numeric(ADK_VS_DH$Metric) >1), ]$Genes
gene_DI <- ADK_VS_DI[which(ADK_VS_DI$`Pvalue anova`<0.05 & as.numeric(ADK_VS_DI$Metric) >1), ]$Genes

diag <- venn.diagram(x = list(
  A = gene_DHDI,
  B = gene_DI,
  C = gene_DH
  ), NULL, col = "transparent",
  fill = c("cornflowerblue", "darkorchid1", "yellow"),
  alpha = 0.50,
  cex = 1,
  fontfamily = "serif",
  fontface = "bold",
  category.names = c("" , "" , ""),
  cat.col = c("darkblue", "darkgreen", "orange"),
  cat.cex = 1.5,
  cat.pos = 0,
  cat.dist = 0.07,
  cat.fontfamily = "serif",
#rotation.degree = 270,
  margin = 0.2,
 )

cols <- c("cornflowerblue", "darkorchid1", "yellow")
lg <- legendGrob(labels=c("ADK vs DH&DI" , "ADK vs DI" , "ADK vs DH"  ), pch=rep(19,length(c("A","B"))),
                 gp=gpar(col=cols, fill="gray"),
                 byrow=TRUE)

g <- gTree(children = gList(diag))
gridExtra::grid.arrange(g, lg, ncol = 2, widths = c(4,1))

```

## Kruskal-Wallis Rank Sum Test
```{r}
gene_DHDI <- ADK_VS_DHDI[which(ADK_VS_DHDI$`Pvalue Kruskal`<0.05 & as.numeric(ADK_VS_DHDI$Metric) >1), ]$Genes
gene_DH <- ADK_VS_DH[which(ADK_VS_DH$`Pvalue Kruskal`<0.05 & as.numeric(ADK_VS_DH$Metric) >1), ]$Genes
gene_DI <- ADK_VS_DI[which(ADK_VS_DI$`Pvalue Kruskal`<0.05 & as.numeric(ADK_VS_DI$Metric) >1), ]$Genes

diag <- venn.diagram(x = list(
  A = gene_DHDI,
  B = gene_DI,
  C = gene_DH
  ), NULL, col = "transparent",
  fill = c("cornflowerblue", "darkorchid1", "yellow"),
  alpha = 0.50,
  cex = 1,
  fontfamily = "serif",
  fontface = "bold",
  category.names = c("" , "" , ""),
  cat.col = c("darkblue", "darkgreen", "orange"),
  cat.cex = 1.5,
  cat.pos = 0,
  cat.dist = 0.07,
  cat.fontfamily = "serif",
#rotation.degree = 270,
  margin = 0.2,
 )

cols <- c("cornflowerblue", "darkorchid1", "yellow")
lg <- legendGrob(labels=c("ADK vs DH&DI" , "ADK vs DI" , "ADK vs DH" ), pch=rep(19,length(c("A","B"))),
                 gp=gpar(col=cols, fill="gray"),
                 byrow=TRUE)

g <- gTree(children = gList(diag))
gridExtra::grid.arrange(g, lg, ncol = 2, widths = c(4,1))
```

